Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.tex
Missing only some parts of the methods and discussion

\hypertarget{overview-of-sections}{%
\section{Overview of sections}\label{overview-of-sections}}

Intro

\begin{itemize}
\item
  This is missingness
\item
  This is how we solve missingness
\item
  For that we need an algorithm, but what if it has not converged?
\item
  That is what we solve in this paper
\end{itemize}

Algorithmic convergence

\begin{itemize}
\item
  What is algorithmic convergence?
\item
  When is it converged? Stationarity versus mixing.
\item
  How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.
\end{itemize}

Methods

\begin{itemize}
\item
  Which methods are there? And how do they relate to mixing and stationarity?
\item
  \(\widehat{R}\)
\item
  AC
\item
  We can use these on any \(\theta\)
\end{itemize}

The Simulation study, Results, Discussion.

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, calculations are not defined, \textbf{hence} statistical models cannot do fitted to the data. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}. An alternative is to apply imputation. With imputation, we `fill in' the missing values in an incomplete dataset. Subsequently, the model of scientific interest can be fitted to the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part \(y_{\rm mis}\) of an incomplete dataset is imputed \(m\) times. This creates \(m\) sets of imputed data \(\dot y_{\rm imp, \ell}\), where \(\ell = 1, 2, ..., m\). The imputed data is then combined with the observed data \(y_{\rm obs}\) to create \(m\) completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate \(Q\): the quantity of scientific interest (e.g., a regression coefficient). Since \(Q\) is estimated on each completed dataset, \(m\) separate \(\hat{Q}_\ell\)-values are obtained. Finally, the \(m\) \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\). The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---inference \citep{rubin96}.

@@

\caption{Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values \(\dot y_{\rm imp, \ell}\) are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\). The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}. \textbf{Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.} Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number. Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}. The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}). We translate the results of the study into guidelines for applied researchers. \textbf{{[}add more societal relevance: what is the consequence of non-convergence? see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations. Chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.

\textbf{{[}SEGWAY to this section{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\). We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\), plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. \textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to bias in the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently also to the pooled estimate \(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with section below{]}}. Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\). Ideally, a multivariate \(\theta\) should be monitored. A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

\textbf{{[}MERGE with section above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (\(T\geq2\)), and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

\begin{align*}
W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
\end{align*}

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain. \textbf{{[}Note that this is not the usual B in MI{]}}

@@

\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which approximates the total variance of \(\theta\). \(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{equation*}

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence.

@@

Following the same notation, we define autocorrelation as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain.

@@

We can interpret \(AC\)-values as a measure of stationarity. \(AC\)-values close to zero indicate independence between subsequent \(\theta\)s in imputation chains. Negative \(AC\)-values show no threat to the stationarity component of convergence. On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence. As convergence diagnostic, the interest is therefore in positive \(AC\)-values. Positive \(AC\)-values occur when \(\theta\)-values of subsequent iterations are similar. When high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values, we speak of recurrence. Recurrence within imputation chains indicates trending. Ergo, non-stationarity may be diagnosed when \(AC\) values are larger than zero.

@@

It is unlikely that an MCMC algorithm will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\). Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is some dependency between subsequent iterations of imputation chains (\(AC > 0\)). In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold. The conventional threshold to diagnose non-mixing is \(\widehat{R} > 1.2\) \citep{gelm92}. Another generally accepted threshold is \(\widehat{R} > 1.1\) \citep{gelm13}. And more recently, \citet{veht19} proposed an even more stringent threshold of \(\widehat{R} > 1.01\). The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}. \(AC\)-values that are significantly higher than zero indicate non-stationarity. \textbf{{[}think about one-sided test{]}}

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But \(\widehat{R}\) and \(AC\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state. To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

@@

When we look at panel B, we conclude something weird. The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter. Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on \texttt{stats::acf()} \citep{R}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms. The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower. Therefore, we compute \(AC\) manually, see the solid line. These \(AC\) values indicate non-stationarity as expected. We therefore calculate \(AC\) manually.

>>>>

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}. An alternative is to apply imputation. With imputation, we `fill in' the missing values in an incomplete dataset. Subsequently, the model of scientific interest can be fitted to the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part \(y_{\rm mis}\) of an incomplete dataset is imputed \(m\) times. This creates \(m\) sets of imputed data \(\dot y_{\rm imp, \ell}\), where \(\ell = 1, 2, ..., m\). The imputed data is then combined with the observed data \(y_{\rm obs}\) to create \(m\) completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate \(Q\): the quantity of scientific interest (e.g., a regression coefficient). Since \(Q\) is estimated on each completed dataset, \(m\) separate \(\hat{Q}_\ell\)-values are obtained. Finally, the \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\). The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand \(Q\) \citep{rubin96}.

@@

\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations \(\dot y_{\rm imp, \ell}\) is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\). The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. \textbf{Therefore, algorithmic convergence should be monitored carefully, }although not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to formulate an informed advice on when approxiamte convergence may be safely concluded. We will therefore assess the performance of different methods to identify non-convergence. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., \textbf{the average of the imputed values}). The collection of \(\theta\)-values between the \textbf{initial value} (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains. If one of these two requirements is not met, there is non-convergence in the algorithm. Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. The distribution of imputed values then differs across imputations. This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)). \textbf{In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\). Ergo, non-convergence leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with sections below{]}} Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number. Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.

\textbf{{[}MERGE with section above and below{]}} Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\). Ideally, a multivariate \(\theta\) should be monitored. A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest. For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

\begin{align*}
W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
\end{align*}

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{{[}Note that this is not the usual B in MI{]}}

@@

\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which \textbf{approximates} the total variance of \(\theta\). \(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
\end{equation*}

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. The assumption underlying this interpretation is that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains \(B\) is then equivalent to the variance within chains \(W\), and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence.

@@

Autocorrelation is defined as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain. Following the same notation as for \(\widehat{R}\),

@@

We can interpret \(AC\)-values as a measure of non-stationarity. If there is dependence between subsequent \(\theta\)-values in imputation chains, \(AC\)-values are non-zero. Positive \(AC\)-values occur when \(\theta\)-values are recurring (i.e., high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values). Recurrence within imputation chains may lead to trending. Negative \(AC\)-values occur when \(\theta\)-values of subsequent iterations are less similar, or diverge from one another. Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence. \textbf{Complete stationarity is reached when \(AC=0\).} As non-convergence diagnostic, our interest is in positive \(AC\)-values.

@@

It is unlikely that iterative algorithms such as MICE will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\). Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim is therefore to reach approximate convergence \citep{gelm13}. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is little dependency between subsequent iterations of imputation chains (\(AC > 0\)). \textbf{Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold. {[}or{]} In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.} The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}. \citet{veht19} proposed a much more stringent threshold of \(\widehat{R} > 1.01\). The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}. \(AC\)-values that are significantly higher than zero indicate non-stationarity. \textbf{{[}think about one-sided test{]}}

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}. The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

@@

\textbf{{[}REPHRASE as active about panel A{]}} From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). In this scenario, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
